Gompertz growth as infinitely correlated non-local growth

### 2023-09-18 arxiv2/new_draft.tex
minor fixes

% \title{On the nature of the Gompertz in relation to the epidemic in 2019-2021}
\title{Gompertzian growth as infinitely correlated non-local growth}

@@

\begin{abstract}
Logistic and Gompertzian growth has traditionally been connected by augmenting the logistic model with an extra parameter to obtain what is known as $\theta$-logistic growth or the Richards model for growth. In spite of this bridge, the biological foundation of Gompertzian growth is only vaguely understood. To add to our biological understanding, we show that there is another way to connect the two growth modes by adding higher order terms in the growth equation, and present both a macroscopic and microscopic representation of growth where the latter involves a network of entities. The microscopic representation elucidates the source of Gompertz growth as the potential abundance and not the realized abundance, and an example of such a network is presented. For finite higher order terms we show that the growth equation can be represented by Gauss' Hypergeometric function, and only for infinitely many terms does the abundance lend itself conveniently to a log-transformation wherein the log-abundance growth velocity decreases linearly, the trademark of Gompertzian growth. We thus equate Gompertzian growth with correlated microscopic particles that exhibit long-range or non-local communication. This conclusion adds to the interpretation of biological systems that undergo Gompertzian growth where long-range communication could be interpreted as the source of growth being external to the individual particles, like a field that stimulates growth simultaneously across all entities. In situations of Gompertzian decay this field could be interpreted as an environmental stressor like air-pollution, age, radiation, etc, while in situations of Gompertzian growth the field could be an internal or external electromagnetic field. Our new discovery allows for a unification of well-known growth models as follows: time independent and uncorrelated growth (or decay) yields the exponential growth function, time dependent and uncorrelated growth yields Gaussian growth, time independent and pairwise correlated growth yields the logistic function and time independent and infinitely correlated growth yields the Gompertz function.

@@

In 1825, Gompertz published a paper on the nature of mortality where he discovered that from adulthood and onward ``the power to avoid destruction'' decreases exponentially with age. Equivalently, mortality rates increase in geometric progression for arithmetic increases in age, an observation now commonly referred to as Gompertz' law of mortality \cite{Gompertz1825}. Makeham (1889) later provided an interpretation to Gompertz' law where the recuperative force of a human, or ``vital force'', becomes less efficient with time \cite{makeham1889further}.

This simple law has been shown to generalize to situations where mortality rates are substituted by many types of abundance variables, like the growth of animals, plants, bacteria, and cancer tumors \cite{weymouth1931relative,weymouth1931age,laird1964dynamics,zwietering1990modeling,skinner1994mathematical,starck1998avian,aggrey2002comparison,paine2012fit,benzekry2014classical,halmi2014evaluation,tjorve2010shapes}, exposing Gompertz growth as ubiquitous in nature. The idea that the Gompertz model can be applied to other biological systems was first proposed by Weymouth (1931) \cite{weymouth1931relative,weymouth1931age} and Winsor (1932) \cite{Winsor1932}. Gompertz law has also been shown to apply to societal phenomena like growth in railway traffic, and the demand for goods and services, and sales of tobacco \cite{olshansky1997ever,prescott1922law,peabody1924growth}.

Alongside the many contexts wherein Gompertz' law has been observed, an almost equal number of efforts have been made to interpret and derive this law from various phenomenological principles \cite{bajzer1997mathematical}. This effort has often been accompanied by a phenomenological comparison of another popular growth model called logistic growth, or Verhulst growth, where ``the power to avoid destruction'' decreases linearly with time, as opposed to decreasing exponentially with time as seen with the Gompertz model. In a study comparing the Gompertz and the logistic growth models, Petroni et al. (2020) \cite{petroni2020logistic} show that Gompertz growth can be interpreted as maximally coherent growth based on long-range communication or correlation, while logistic growth is minimally coherent without long-range behavior. With this interpretation, one is lead to postulate that the Gompertz function could be a result of a global or external stimuli while the logistic function is a result of internal or local interaction. As such, the logistic function has been popular both for communicable disease models \cite{harko2014exact,kroger2020analytical,schlickeiser2021analytical,heng2020approximately} and and for predator-prey theory \cite{berryman1992orgins}, where disease or injury spreads through pairwise interactions.

A number of studies in the biological context have argued that the Gompertzian model could be sourced or activated from an external factor which is ubiquitously and instantaneously introduced to the system, e.g. as a toxic agent or an environmental stressor to which the system gradually adapts \cite{neafsey1988gompertz,neafsey1989gompertz,thompson1990biphasic}. In the original case of mortality rates as a function of age, this interpretation would imply that time itself is the environmental ubiquitous stressor. Other environmental stressors like chronic radiation exposure have been observed to shift the Gompertz curve, an observation also theoretically developed by Sacher (1956) (\cite{sacher1956statistical}). Sacher modeled the toxic stressor as a stochastic perturbation to the physiologic state of a population wherein beyond a certain point of perturbation in a given entity would result in its death. The interpretation of the Gompertz curve as the result of a ubiquitous stochastic perturbation was recently revived by De Lauro et al. (2014) \cite{de2014stochastic} who arrived at the same result without alluding to a threshold value beyond which death was inevitable, but rather by letting the stochastic perturbation be the source of growth or decay.

Another framework used to show how the Gompertz model can result from an external or ubiquitous stressor is chemical reaction theory. In reaction theory, a set of reactants combine to produce products, possibly in the presence of a catalyst. Morkov (2019) \cite{markov2019reaction} showed that the Gompertz can be interpreted as growth under a catalyst whose effect diminishes at a rate independent of the growth rate of the reactant. In other words, the catalyst can be considered as an external and ubiquitous catalyst to the reactants, perfectly in line with the ubiquitous stochastic stressor model from Sacher and De Lauro et al. On the other hand, if the logistic growth equation is recast into the reaction network equivalent, the catalyst is diminished by the growth of the reactant, which suggests that the source of growth is internal or at least dependent on the internal state. So even in reaction theory the Gompertz model could be seen as a result of a ubiquitous field while the logistic model could be seen as a result of internal interactions.

In this paper, we will elaborate on the idea that the logistic model results from local interactions while the Gompertz model results from a field or equivalently from an infinite set of higher order interactions between the microscopic entities involved in the growth. We will devise a slightly different parametrization than the Richard's model, the generalized Gompertz-Logistic equation which is commonly used to compare the two models (e.g. by \citet{petroni2020logistic}, \citet{tjorve2017use}, and \citet{wang2012richards}). In particular, instead of considering a real number as the unifying exponent we will appeal to integer exponents with a number of terms that tend towards infinity as we approach the Gompertz model. As we shall see, this is rooted in the microscopic interpretation of the growth system, where each subgroup of $n$ elementary entities interacts with all other equally sized subgroups each composed of mutually exclusive elementary entities. If these subgroups are present at all orders of cardinality, the Gompertz model emerges, while only pairwise interactions yield the logistic model. As seen in the aforementioned references, this is in line both with experimental and theoretical results. In presenting this argument, we present a novel bridge between the logistic and the Gompertz model which offers both a mathematical and biological foundation to the source of growth.

It is worth noting that the notion that Gompertz growth is maximally coherent and non-local has also been derived from quantum mechanical principles \cite{molski2003coherent}. Molski and Konarski (2003) showed that the Gompertz equation is the solution of a form of the time-independent Schrodinger equation. From this result, they derive the quantum coherent states which are non-local in space while moving along a classical time trajectory. This adds to our exhibits that the Gompertz model can be interpreted as non-local in origin. Coherent quantum states was originally formulated in the context of electromagnetic fields by \citet{glauber1963coherent} (for which he received the Nobel Prize). Perhaps some of the biological macroscopic growth phenomena references are also due to such fields. For example, the role of internal electrical fields in regenerative growth is well established and is based on negative electrical potentials that act on sensitive cells \cite{becker1984electromagnetic}. The limb-regeneration process of the salamander is an example of such a phenomena.

In fact, it has been shown that cancer malignancies are mostly negative in electromagnetic polarity with respect to the rest of the body, while nonmalignant tumor pathology is mostly positive \cite{langman1949technique}. One is then led to the interpretation that certain biological growth processes are coupled with an electromagnetic field of specific character. Consequently and unsurprisingly, the artificial disruption of such a field has been found to disrupt the mitotic cycle or cell cycle \cite{goodman1983pulsing,liboff1981alternating,norton1984bioelectric,liboff1984time}. This disruptive effect also extends to other biological processes, as shown by Wever and Brown's experiments on biological endogenous cycles lie the circadian cycle through manipulation of the environmental electromagnetic field \cite{brown1976biological,wever1974elf}.

Finally, on the topic of disease modelling, the Gompertz model has been shown by Wang et al. \cite{wang2012richards} to not originate from a communicable disease paradigm where only local disease spread is possible. They used the common Susceptible-Infected-Recovered model by Kermack and McKendrick \cite{kermack1927contribution} augmented by the Richards parameter and showed that the allowed parameter range with the communicable disease assumption does not cover the parameter value taken in the Gompertz scenario. This would imply that any mortality patterns that shows a Gompertzian characteristic cannot be a result of a communicable disease, but rather from a ubiquitous stressor.

Our paper is structured as follows: we will first present a new macroscopic foundation of the Gompertz-Logistic spectrum which will offer an alternative to the traditional Richard's model. Then, we present a microscopic foundation which acts as the building blocks of the macroscopic phenomena. This is where we show that the Gompertz model results from an infinite set of higher order interactions between the elementary entities involved in the growth, which could alternatively be seen as a ubiquitous field or in the biological context: an environmental stressor. We proceed with a discussion that summarizes our findings in the context of the other well-known growth modalities involving the Gaussian and the Laplacian model and finally end with a conclusion.

@@

First consider a system that grows according to logistic growth. If we let the variable $X(t)$ denote the abundance of a quantity in a normalized system where its maximum size is 1, the equation that determines its size as a function of time is
\begin{equation}
  \frac{1}{X}\frac{d}{dt} X(t) = \beta (1 - X(t)).
\end{equation}
This is commonly interpreted as the right hand side representing ``resources available for growth'' which decreases linearly with respect to the relative abundance's growth rate on the left hand side. One way to generalize logistic growth is to introduce a parameter that exponentiates the resource availability in the following way,
\begin{equation}
\label{eq:Rich}
  \frac{1}{X}\frac{d}{dt} X(t) = \beta (1 - X^{\theta}(t)),
\end{equation}
also called $\theta$-logistic growth or Richards' model. Petroni et al. (2020) argued that this exponentiation can be interpreted as non-linear interaction effects where there is a level of co-operation in the microscopic domain. They showed that in the limit where the system becomes \textit{maximaly cooperative} or \textit{coherent}, $\theta\rightarrow 0$ and the $\theta$-logistic equation reduces in this limit to the Gompertz growth model,
 \begin{equation}
 \label{eq:Gomp}
  \frac{1}{X}\frac{d}{dt} X(t) = - \beta \ln X(t).
\end{equation}
In this limit, the relative growth goes to infinity as the system approaches $t=0$, which could be seen as a verification of the maximally coherent characteristic of the system. One is reminded of a system that is perturbed and left to reach a new equilibrium. Petroni's definition of maximal coherence in the context of the Gompertz equation was shown to rigorously correspond to the definition of coherence used in quantum mechanical systems \cite{molski2003coherent} attributed to Glauber (1963) \cite{glauber1963coherent}.

@@

Now, consider a network between these $N$ entities along a graph with a set of edges and nodes which delineates the interaction of the growth variables. Then consider the growth equation with only pairwise interactions, or \textit{first order} interactions, governed by an adjacency matrix ${a_{ij}}$ with a binary matrix with ones where the $i^{th}$ and $j^{th}$ nodes are connected, and zeroes otherwise, viz.
\begin{equation}
\label{eq:modSIR}
\frac{d x_i}{dt} = x_i\sum_{j \in \mathcal{N}(i)}{a_{ij}}s_{j} \quad \forall i,
\end{equation}
where $\mathcal{N}(i)$ is the neighborhood of nodes connected to node $i$.
First note that for a single pair this would amount to traditional logistic growth. We choose to work in a unitless domain without parameters for the interaction strength and carrying capacity for simplicity of exposition. These can be added without changing the conclusions of these arguments.

@@

\subsection{Microscopic Gompertz Growth}

@@

We finally note that for the right hand side to exhibit asymptotically infinite growth at the onset of growth, it is not only sufficient but also necessary to have higher order terms. This is seen by virtue of the infinite polynomial representation of the logarithm and implies that Gompertz growth cannot result from pairwise interactions under the assumptions presented in this model, but rather requires all higher order interactions to be present.

@@

As a side note, the Poisson distribution can be included in this discussion as the distribution which governs the infinitesimal growth rate of a coherent or infinitely correlated group of entities. By letting the growth rate be the rate at which a zero-count is observed with the Poisson mean parameter equal to $-\beta t$, one recovers the Gompertz model in Equation \ref{eq:Gomp}. This argument was put forth by \citet{shklovskii2005simple}, and \citet{glauber1963coherent} was the first to point out the connection where the Poisson distribution reflects a coherent system. One the other hand, Glauber also showed that incoherent systems, or systems with independent entities, follow the Gaussian distribution which is in accordance with our results and discussion here. The mapping between coherence and incoherence has also been postulated as a medical diagnostic technique by \citet{zhang1994log}, where incoherence is a sign of disease while coherence is a sign of well-being.

\section{Conclusion}
\label{sec:discussion}

@@

An infinitely interacting system where entities are classical in nature with well-defined position and momentum, breaks the law of causality unless we allow the source of growth to be ubiquitous field, externally or internally created, which affects all microscopic entities instantaneously \footnote{Or at speeds much greater than the differential times in the growth equations}. With a such a source, entities need not be interacting with each other but rather respond to the field-induced stimuli. Examples of such models are already given by De Lauro (2014), Sacher (1956) and Markov (2019). \citet{de2014stochastic} and \citet{sacher1956statistical} consider the source of growth to be stochastic in nature, and while Lauro et al. do not discuss the external nature of such a source, their model does not preclude it. More generally, with growth as stochastically rooted it is also ubiquitously present uniformly across all entities from the very onset of growth. Markov (2019)\cite{markov2019reaction} discusses growth within the context of chemical reactions, where the source of growth is a catalytic agent that is implicitly available ubiquitously to all reactants to the same degree, exactly in line with De Lauro and Sacher's models.

@@

It is worth noting that the original formulation of the Gompertz model made by Gompertz himself was concerning mortality rates as a function of age. He noticed that mortality rates increased at a geometric rate with age. In this context the ubiquitous stressor postulated above is of course time itself and thus Gompertz growth in this context will be implicitly \textit{time dependent}. In other words, Gompertz growth arises without the need for communication across entities, but is mediated by a mutual field. One could then see the entities as communicating indirectly though this field, in a perceived one-way fashion from the field to each entity.

@@

\section*{Supplement}

>>>>

% \title{On the nature of the Gompertz in relation to the epidemic in 2019-2021}
\title{Gompertz growth as infinitely correlated non-local growth}

@@

\begin{abstract}
Logistic and Gompertz growth has traditionally been connected by augmenting the logistic model with an extra parameter to obtain what is known as $\theta$-logistic growth or the Richards model for growth. In spite of this bridge, the biological foundation of Gompertz growth is only vaguely understood. To add to our biological understanding, we show that there is another way to connect the two growth modes by adding higher order terms in the growth equation, and present both a macroscopic and microscopic representation of growth where the latter involves a network of entities. The microscopic representation elucidates the source of Gompertz growth as the potential abundance and not the realized abundance, and an example of such a network is presented. For finite higher order terms we show that the growth equation can be represented by Gauss' Hypergeometric function, and only for infinitely many terms does the abundance lend itself conveniently to a log-transformation wherein the log-abundance growth velocity decreases linearly, the trademark of Gompertz growth. We thus equate Gompertz growth with correlated microscopic entities that exhibit long-range or non-local interaction. This conclusion adds to the interpretation of biological systems that undergo Gompertz growth where long-range interaction could be interpreted as the source of growth being external to the individual entities, like a field that stimulates growth simultaneously across all entities. This field could be interpreted as an environmental stressor like air-pollution, age, radiation, etc, or an environmental stimulant like an electromagnetic field. Our new discovery allows for a unification of well-known growth models as follows: time independent and uncorrelated growth (or decay) yields the exponential growth function, time dependent and uncorrelated growth yields Gaussian growth, time independent and pairwise correlated growth yields the logistic function and time independent and infinitely correlated growth yields the Gompertz function.

@@

In 1825, Gompertz published a paper on the nature of mortality where he discovered that from adulthood and onward ``the power to avoid destruction'' decreases exponentially with age. Equivalently, mortality rates increase in geometric progression for arithmetic increases in age, an observation now commonly referred to as Gompertz' law of mortality \cite{Gompertz1825}. Makeham (1889) later provided an interpretation to Gompertz' law where the recuperative force of a human, or ``vital force'', becomes less efficient with time \cite{makeham1889further}. Gompertz' law has later been known as the Gompertz model or Gompertz growth or decay.

\subsection{Gompertz' law in existing work}
This simple law has been shown to generalize to situations where mortality rates are substituted by many types of abundance variables, like the growth of animals, plants, bacteria, and cancer tumors \cite{weymouth1931relative,weymouth1931age,laird1964dynamics,zwietering1990modeling,skinner1994mathematical,starck1998avian,aggrey2002comparison,paine2012fit,benzekry2014classical,halmi2014evaluation,tjorve2010shapes}, exposing Gompertz growth as nearly ubiquitous in biological growth and decay processes. The idea that the Gompertz' law can be applied to other biological systems was first proposed by Weymouth (1931) \cite{weymouth1931relative,weymouth1931age} and Winsor (1932) \cite{Winsor1932}. Gompertz' law has also been shown to apply to societal phenomena like growth in railway traffic, and the demand for goods and services, and sales of tobacco \cite{olshansky1997ever,prescott1922law,peabody1924growth}.

Alongside the many contexts wherein Gompertz' law has been observed, an almost equal number of efforts have been made to interpret and derive this law from various phenomenological principles \cite{bajzer1997mathematical}. This effort has often been accompanied by a phenomenological comparison of another popular growth model called logistic growth, or Verhulst growth, where ``the power to avoid destruction'' decreases linearly with time, as opposed to decreasing exponentially with time as seen with the Gompertz model. In a study comparing the Gompertz and the logistic growth models, Petroni et al. (2020) \cite{petroni2020logistic} show that Gompertz growth can be interpreted as \textit{maximally coherent} growth based on long-range interaction or correlation, while logistic growth is minimally coherent without long-range behavior. With this interpretation, one is lead to postulate that the Gompertz model is a result of an external stimulus with perceived long-range interactions while the logistic model is a result of an internal stimulus with local interaction. As such, the logistic model has been popular both for communicable disease models \cite{harko2014exact,kroger2020analytical,schlickeiser2021analytical,heng2020approximately} and and for predator-prey theory \cite{berryman1992orgins}, where disease or injury spreads through pairwise interactions.

A number of studies in the biological context have argued that Gompertz growth could be activated from an external factor which is ubiquitously and instantaneously introduced to the system, e.g. as a toxic agent or an environmental stressor to which the system gradually adapts \cite{neafsey1988gompertz,neafsey1989gompertz,thompson1990biphasic}. In the original case of mortality rates as a function of age, this interpretation would imply that simply living on earth is the ubiquitous stressor. Other environmental stressors like chronic radiation exposure have been observed to shift the Gompertz curve, an observation also theoretically developed by Sacher (1956) (\cite{sacher1956statistical}). Sacher modeled the toxic stressor as a stochastic perturbation to the physiologic state of a population wherein death of an entity would occur beyond some perturbation threshold. The interpretation of Gompertz growth as the result of a ubiquitous stochastic perturbation was recently revived by De Lauro et al. (2014) \cite{de2014stochastic} who arrived at the same result without alluding to a threshold value beyond which death was inevitable, but rather by letting the stochastic perturbation be the source of growth or decay.

Another framework used to show how the Gompertz model can result from an external or ubiquitous stressor is chemical reaction theory. In reaction theory, a set of reactants combine to produce products, possibly in the presence of a catalyst. Morkov (2019) \cite{markov2019reaction} showed that the Gompertz model can be interpreted as growth under a catalyst whose effect diminishes at a rate independent of the growth rate of the reactant. In other words, the catalyst can be considered as an external and ubiquitous stimulant or stressor to the reactants, perfectly in line with the ubiquitous stochastic perturbation model from Sacher and De Lauro et al. On the other hand, if the logistic growth equation is recast into the reaction network equivalent, the catalyst is diminished by the growth of the reactant, which suggests that the source of growth is internal or at least dependent on the internal state. So even in reaction theory the Gompertz model could be seen as a result of a ubiquitous field while the logistic model could be seen as a result of internal interactions of the reactants.

It is worth noting that the notion that Gompertz growth is maximally coherent and non-local has also been derived from quantum mechanical principles \cite{molski2003coherent}. Molski and Konarski (2003) showed that the Gompertz equation is the solution of a form of the time-independent Schrodinger equation. From this result, they derive the quantum coherent states which are non-local in space while moving along a classical time trajectory. This adds to our exhibits that the Gompertz model can be interpreted as non-local in origin. Coherent quantum states was originally formulated in the context of electromagnetic fields by \citet{glauber1963coherent} (for which he received the Nobel Prize). Perhaps some of the biological macroscopic growth phenomena references are also due to such fields. For example, the role of internal electrical fields in regenerative growth is well established and is based on negative electrical potentials that act on sensitive cells \cite{becker1984electromagnetic}.

In fact, it has been shown that cancer malignancies are mostly \textit{negative} in electric polarity with respect to the rest of the body, while nonmalignant (or benign) tumor pathology is mostly \textit{positive} in polarity \cite{langman1949technique}. Gompertz growth has also been widely observed in cancerous cells \cite{kozusko2003combining,laird1964dynamics}. One is then led to the interpretation that certain biological growth processes that exhibit Gompertz growth are coupled with an electromagnetic field of specific character. Consequently and unsurprisingly, the artificial disruption of such a field has been found to disrupt the mitotic cell cycle \cite{goodman1983pulsing,liboff1981alternating,norton1984bioelectric,liboff1984time}.

Finally, on the topic of disease modelling, the Gompertz model has been shown by Wang et al. \cite{wang2012richards} to be excluded from a communicable disease paradigm where only local disease spread is possible. They used the common Susceptible-Infected-Recovered (SIR) model by \citet{kermack1927contribution} augmented by the Richards parameter and showed that the allowed parameter range with the communicable disease assumption does not cover the parameter value taken in the Gompertz scenario. Given this observation by Wang et al., any mortality pattern that shows a Gompertz characteristic cannot be a result of a communicable disease, but rather from a ubiquitous stressor. While many studies show the SIR model (and its extensions) as the fitted model in epidemics \cite{carletti2020covid,cooper2020sir,postnikov2020estimation,munoz2021sir,cooper2022dynamical,saikia2021covid}, some recent studies have successfully fitted a Gompertz model \cite{Ohnishi2020,Rypdal2020,Catala2020,rodrigues2020monte,Levitt2020}.

\subsection{Main Results and Roadmap}
In this paper, we will elaborate on the idea that the logistic model results from local interactions while the Gompertz model results from a field or equivalently from an infinite set of higher order interactions between the microscopic entities involved in the growth. We will devise a slightly different parametrization than the Richard's model, the generalized Gompertz-Logistic equation which is commonly used to compare the two models (e.g. by \citet{petroni2020logistic}, \citet{tjorve2017use}, and \citet{wang2012richards}). In particular, instead of considering a decimal number as the unifying exponent we will appeal to integer exponents with a number of terms that tend towards infinity as we approach the Gompertz model. As we shall see, this is rooted in the microscopic interpretation of the growth system, where each subgroup of $n$ elementary entities interacts with all other equally sized subgroups each composed of mutually exclusive elementary entities. If these subgroups are present at all orders of cardinality, the Gompertz model emerges, while only pairwise interactions yield the logistic model. As seen in the aforementioned references, this is in line both with experimental and theoretical results. In presenting this argument, we present a novel bridge between the logistic and the Gompertz model which offers both a mathematical and biological foundation to the source of growth.

Our paper is structured as follows: we will first present a new macroscopic foundation of the Gompertz-Logistic spectrum which will offer an alternative to the traditional Richard's model. Then, we present a microscopic foundation which acts as the building blocks of the macroscopic phenomena. This is where we show that the Gompertz model results from an infinite set of higher order interactions between the elementary entities involved in the growth, which could alternatively be seen as a ubiquitous field or in the biological context: an environmental stressor. We proceed with a discussion that summarizes our findings in the context of the other well-known growth modalities involving the Gaussian and the Laplacian model and then conclude.

@@

First consider a system that grows according to logistic growth. If we let the variable $X(t)$ denote the abundance of a quantity in a normalized system where its maximum size is 1, the equation that determines its size as a function of time is
\begin{equation}
  \frac{1}{X}\frac{d}{dt} X(t) = \beta (1 - X(t)).
\end{equation}
This is commonly interpreted as the right hand side representing ``resources available for growth'' which decreases linearly with respect to the relative abundance's growth rate on the left hand side. One way to generalize logistic growth is to introduce a parameter that exponentiates the resource availability in the following way,
\begin{equation}
\label{eq:Rich}
  \frac{1}{X}\frac{d}{dt} X(t) = \beta (1 - X^{\theta}(t)),
\end{equation}
also called $\theta$-logistic growth or Richards' model. Petroni et al. (2020) argued that this exponentiation can be interpreted as non-linear interaction effects where there is a level of cooperation in the microscopic domain. They showed that in the limit where the system becomes \textit{maximaly cooperative} or \textit{coherent}, $\theta\rightarrow 0$ and the $\theta$-logistic equation reduces in this limit to the Gompertz growth model,
 \begin{equation}
 \label{eq:Gomp}
  \frac{1}{X}\frac{d}{dt} X(t) = - \beta \ln X(t).
\end{equation}
In this limit, the relative growth goes to infinity as the system approaches $t=0$, which could be seen as a verification of the maximally coherent characteristic of the system. One is reminded of a system that is perturbed and left to reach a new equilibrium. Petroni's definition of maximal coherence in the context of the Gompertz equation was shown to rigorously correspond to the definition of coherence used in quantum mechanical systems \cite{molski2003coherent} attributed to Glauber (1963) \cite{glauber1963coherent}.

@@

Now, consider a network between these $N$ entities along a graph with a set of edges and nodes which delineates the interaction of the growth variables, where we define interaction in this context as the transfer of one entity's potential growth to another entity which converts potential to realized growth. Then consider the growth equation with only pairwise interactions, or \textit{first order} interactions, governed by an adjacency matrix ${a_{ij}}$ with a binary matrix with ones where the $i^{th}$ and $j^{th}$ nodes are connected, and zeroes otherwise, viz.
\begin{equation}
\label{eq:modSIR}
\frac{d x_i}{dt} = x_i\sum_{j \in \mathcal{N}(i)}{a_{ij}}s_{j} \quad \forall i,
\end{equation}
where $\mathcal{N}(i)$ is the neighborhood of nodes connected to node $i$.
First note that for a single pair this would amount to traditional logistic growth. We choose to work in a unitless domain without parameters for the interaction strength and carrying capacity for simplicity of exposition. These can be added without changing the conclusions of these arguments.

@@

\subsection{Microscopic Gompertz Growth}
\label{sec:microGomp}

@@

We finally note that for the right hand side to exhibit asymptotically infinite growth rate at the onset of growth, it is not only sufficient but also necessary to have higher order terms. This is seen by virtue of the infinite polynomial representation of the logarithm and implies that Gompertz growth cannot result from pairwise interactions under the assumptions presented in this model, but rather requires all higher order interactions to be present.

@@

As a side note, the Poisson distribution can be included in this discussion as the distribution which governs the infinitesimal growth rate of a coherent or infinitely correlated group of entities. By letting the growth rate be the rate at which a zero-count is observed with the Poisson mean parameter equal to $-\beta t$, one recovers the Gompertz model in Equation \ref{eq:Gomp}. This argument was put forth by \citet{shklovskii2005simple}, and \citet{glauber1963coherent} was the first to point out the connection where the Poisson distribution reflects a coherent system. Glauber also showed that incoherent systems, or systems with independent entities, follow the Gaussian distribution which is in accordance with our results and discussion here. The mapping between coherence and incoherence has also been postulated as a medical diagnostic technique by \citet{zhang1994log}, where incoherence is a sign of disease while coherence is a sign of well-being.

\section{Discussion}
\label{sec:discussion}

@@

An infinitely interacting system where entities are classical in nature with well-defined position and momentum, breaks the law of causality unless we allow the source of growth to be a ubiquitous field, externally or internally created, which affects all microscopic entities instantaneously \footnote{Or at speeds much greater than the differential times in the growth equations}. With a such a source, entities need not be interacting with each other but rather respond to the field-induced stimuli. Examples of such models are already given by De Lauro (2014), Sacher (1956) and Markov (2019). \citet{de2014stochastic} and \citet{sacher1956statistical} consider the source of growth to be stochastic in nature, and while Lauro et al. do not discuss the external nature of such a source, their model does not preclude it. More generally, with growth as stochastically rooted it is also ubiquitously present uniformly across all entities from the very onset of growth. Markov (2019)\cite{markov2019reaction} discusses growth within the context of chemical reactions, where the source of growth is a catalytic agent that is implicitly available ubiquitously to all reactants to the same degree, exactly in line with De Lauro and Sacher's models.

@@

It is worth noting that the original formulation of the Gompertz model made by Gompertz himself was concerning mortality rates as a function of age. He noticed that mortality rates increased at a geometric rate with age. In this context the ubiquitous stressor postulated above would be the living conditions on our planet, for which time acts as a proxy. One could further postulate this stressor as a field surrounding the earth, e.g. in its electromagnetic footprint often related with the Schumann Resonance \cite{cherry2002schumann,schumann1952strahlungslosen}. It is indeed curious that we do not see mortality rates behave as linearly dependent on time, for which a Gaussian curve would be appropriate. Nor do we see any finite-ordered polynomial time dependence on mortality rates. Instead we see growth of mortality rates as sourced from other entities at all interaction orders, which in light of our discussion above is equivalent to mortality coming from an environmental stressor.

@@

\section*{Appendix A}
Here we arrive at a closed-form expression for a finite number of interaction terms in this modified logistic model, which represents the infection term in the SIR model. Working with a single unitless growth variable and omitting the multivariate network without loss of generality, we start with the logistic growth equation augmented by $K$ terms and using the nomenclature defined in Section \ref{sec:micro}, viz.

\begin{equation}
\frac{d \ln{x}}{dt} = \sum_{i=1}^K s^i/i.
\end{equation}
Using standard integrals we see that this sum amounts to a Gompertz growth term adjusted by a Hypergeometric function,
\begin{eqnarray*}
  \sum_{i=1}^K s^i/i &=& \int_s \sum_{i=0}^{K-1} s^i ds\\
  &=&\int_s \frac{1-s^K}{1-s} ds\\
  &=&- \frac{{}_{1}F_{2}({1,K+1,K+2;s})}{K+1}s^{K+1} - \ln(1-s),
\end{eqnarray*}
where ${}_1F_{2}$ is Gauss' Hypergeometric function, and integration limits are indefinite. In summary, the modified logistic function emerges as
 \begin{equation}
 \label{eq:modLogistic2}
\frac{d}{dt}\ln{x(t)} = - \frac{{}_{1}F_{2}({1,K+1,K+2;1-x(t)})}{K+1}(1-x(t))^{K+1} - \ln{x(t)},
 \end{equation}
 which represents the novel bridge between logistic and Gompertzian growth. Note that for $K=1$, this relative growth rate becomes the standard logistic function, while increasing the number of higher order terms will move the equation closer to the Gompertz curve as the Hypergeometric term approaches zero. This modification to the logistic function does not have a 1-to-1 correspondence with the Richards model, but serves instead as an alternative when encoding non-linear or collaborative growth effects (see Fig. \ref{fig:hypergeometric_plain}).
